---
title: "Monarchs assignment"
author: "Jonas Møgeltoft Mehl"
date: "21-08-2026"
output: html_document
---

```{r setup, include=TRUE}
library(tidyverse)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the monarchs

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `monarchs.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the separator: semicolon  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
library(tidyverse)

monarchs1 <- read.csv("data/tidied_monarch_data.csv")

monarchs2 <- read_csv("data/tidied_monarch_data.csv")

monarchs3 <- read.csv2("data/tidied_monarch_data.csv")

monarchs4 <- read_csv2( "data/tidied_monarch_data.csv")
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `monarchs` object

read_csv2 saves as a tbl, which is a part of the tidyverse. And it also loads the csv correctly with the semicolon separator

2. What is the result of running `class()` on the `monarchs` object created with a tidyverse function

We see that the class is a tbl data frame.

3. How many columns does the object have when created with these different functions?

11 columns

4. Show the dataset so that we can see how R interprets each column

The names are loaded correctly as characters, and all the dates are also loaded correctly as dates.

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

monarchs <- monarchs4

class(monarchs)


ncol(monarchs)


glimpse(monarchs)


```


## Calculate the duration of reign for all the monarchs in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the monarchs dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE
library(lubridate)
monarchs <- monarchs %>% 
  mutate(duration = as.numeric(year(last_term_end) - year(first_term_start)))
```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `monarchs` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
mean_duration <- mean(monarchs$duration, na.rm = TRUE)
```

## How many and which monarchs enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `monarchs` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
longer_monarchs <- monarchs %>% 
  filter(duration > mean_duration) 
  
longer_monarchs

count(longer_monarchs)
```


## How many days did the three longest-ruling monarchs rule? 

* Sort monarchs by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
monarchs <- monarchs %>% 
  mutate(Days = duration * 365.24) #Average number of days in a year, calculated with leap years.

sliced_monarchs <- monarchs  %>% 
  arrange(desc(duration)) %>% 
  slice_head(n=3)

sliced_monarchs
```


## Challenge: Plot the monarchs' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
#Creating midyear colom

monarchs <- monarchs %>% 
  mutate(midyear = as.numeric(year(first_term_start) + ( year(last_term_end) - year(first_term_start) )/2))


#Creating gg_plot plot object
monarchs_plot <- monarchs %>% 
  ggplot(aes(x= midyear, y = duration)) +
    labs(title = "Duration of reign of Danish monarcs",
         x = "Middle of Reign [Year]",
         y = "Duration of Reign [Years]") +
  theme_bw() +
  theme(text = element_text(size = 14))

#Creating point plot
monarchs_plot +
  geom_point(alpha = 0.5, na.rm = TRUE)  +
  geom_smooth(alpha = 0.5, na.rm = TRUE)
```
By using the geom_smooth together with geom_point one can see the tendency of reignlength becomming longer with time.


And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!